---
title: "EDA - group by and summarize"
author: Tanya Butt
output: github_document
  
---

```{r}
library(tidyverse)
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6, 
  out.width = "90%"
)
```
Load in a dataset that we'll use often.
```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```

## 'group_by'

```{r}
weather_df %>% 
 group_by(name, month)
```

## count some things

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n())

weather_df %>% 
  group_by(name, month) %>% 
  summarize(n_obs = n())

weather_df %>% 
  count(name)

weather_df %>% 
  count(month, name = "n_obs")

```

It's nice that summarize produces a data frame.
'table' doesn't do that, and it's bad.

```{r}
weather_df %>% 
  pull(name) %>% 
  table()
```

```{r}
weather_df %>% 
  janitor::tabyl(month, name)
```

## More general summaries 

Let's count, take means, look at SDs...

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n_obs = n(),
    mean_tmax = mean(tmax, na.rm = TRUE),
    sd_tmax = sd(tmax, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() + 
  geom_line()
```

Formatting table outputs.

```{r}
weather_df %>% 
  group_by(name) %>% 
  summarize(
    n_obs = n(),
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) %>% 
  knitr::kable(digits = 2)
```

## grouped 'mutate'

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```


```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    tmax_rank = min_rank(desc(tmax))
  ) %>% 
  filter(tmax_rank < 2)

weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs = n(),
    n_days = n_distinct(date))

weather_df %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold",
      TRUE ~ ""
    )) %>% 
  filter(name != "Waikiki_HA") %>% 
  group_by(name, cold) %>% 
  summarize(count = n())

weather_df %>% 
  mutate(cold = case_when(
    tmax < 5 ~ "cold",
    tmax >= 5 ~ "not_cold",
    TRUE      ~ ""
  )) %>% 
  filter(name != "Waikiki_HA") %>% 
  janitor::tabyl(name, cold)
```

Lagged Variables 

```{r}
#what's the lag information from the day before
weather_df %>% 
  group_by(name) %>% 
  mutate(
    lagged_tmax = lag(tmax, n = 1),
    tmax_diff = tmax - lagged_tmax
  ) %>% 
  summarize(diff_sd = sd(tmax_diff, na.rm = TRUE))
```

## Limitations 

What if my "summary" is a linear model...

Well...that doesn't work
```{r}
weather_df %>% 
  group_by(name) %>% 
  summarize(cor_tmin_tmax = cor(tmin, tmax, use = "complete"))

weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  lm(tmax ~ tmin, data = .)

```

## general summaries

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax),
    mean_prec = mean(prcp, na.rm = TRUE),
    median_tmax = median(tmax),
    sd_tmax = sd(tmax))
 
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax),
    median_tmax = median(tmax))

weather_df %>% 
  group_by(name, month) %>% 
  summarize(across(tmin:prcp, mean))

weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax)) %>% 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() + geom_line() +
  theme(legend.position = "bottom")

weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax)) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax) %>% 
  knitr::kable(digits = 1)
```

## Grouped mutate

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate
```

